### 2008-03-11 trunk/Papers/dl-gre-08/discussion.tex
wrote first versions of intro, related work, efficiency

\begin{figure}
\includegraphics[width=0.4\textwidth]{drawers.eps}
\caption{A schematic view of the filing cabinets.\todo{needs nicer pic}}\label{fig:drawers}
\end{figure}

@@

>>>>

\begin{figure}
\includegraphics[width=0.4\textwidth]{drawers}
\caption{A schematic view of the filing cabinets.\todo{needs nicer pic}}\label{fig:drawers}
\end{figure}

@@

\todo{should we show some more examples? - ak}

Both the \el\ and the \alc\ algorithms took about 15 milliseconds to
compute distinguishing concepts for all 16 individuals in the Viethen
\& Dale dataset.\footnote{We measured all runtimes on a MacBook Pro
  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta, and allowed the
  Java VM to warm up, i.e.\ just-in-time compile all bytecode, before
  taking the measurements.}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{runtimes}
  \caption{Average runtimes (in ms) of the two algorithms on random graphs of
    different sizes.} 
  \label{fig:runtimes}
\end{figure}

In order to get a more comprehensive picture of our system's
efficiency, we ran the two algorithms on random graphs with increasing
numbers of nodes.  \todo{harmonize this with what Carlos calls them in
  Section~\ref{sec:bisim}} Each graph had ten different atoms and four
different roles, a 10\% chance for a node to have each atom, and a
10\% chance for each pair of nodes to be connected by each role.  The
results (averaged over 10 runs for each graph size) are shown in
Fig.~\ref{fig:runtimes}.  As the figure shows, the \el\ algorithm
takes about 400 ms on average to generate relational REs for all
individuals in the graph with 100 nodes, i.e.\ 4 ms on average for
each individual.  The \alc\ algorithm is slower by a constant factor
of about 2-3 which probably reflects implementation differences.  In
both cases, we think we can conclude that the systems are fast enough
for practical use.

\subsection{Interface to realization}

\begin{itemize}
\item there's a risk to generate a concept that can't be realized,
  especially for the \alc\ algorithm
\item everybody except perhaps SPUD has this problem
\item problem is worse in our case because it is harder to control the
  order in which atoms and relations are explored
\item would be interesting for future work to think about how
  search in the bisim algorithms can be controlled by the realizer
\end{itemize}

### 2008-03-11 trunk/Papers/dl-gre-08/introduction.tex
added some more references
wrote first versions of intro, related work, efficiency

\cite{Dale1995}
\cite{dale91:_gener_refer_expres_invol_relat}

>>>>

The generation of referring expressions is one of the most active and
successful research areas in NLG.  Starting with Dale and Reiter's
incremental algorithm \cite{Dale1995}, various researchers have added
extensions such as reference to sets \cite{Stone2000}, more expressive
logical connectives \cite{deemter01:_gener_refer_expres}, and
relational expressions \cite{dale91:_gener_refer_expres_invol_relat}.

Referring expressions involving relations, in particular, have
received increasing attention recently, particularly in the context of
spatial referring expressions in situated generation (e.g.\
\cite{kelleher06:_increm_gener_of_spatial_refer}), where it seems
particularly natural to use expressions such as ``the book on the
table''.  However, the classical algorithm by
\newcite{dale91:_gener_refer_expres_invol_relat} was recently shown to
be unable to generate satisfying REs in practice
\cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
\& Haddock algorithm and most of its successors
\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} are
vulnerable to the problem of ``infinite regress'', where the algorithm
jumps back and forth between generating descriptions for two related
individuals infinitely, as in ``the book on the table which supports a
book on the table \ldots''.

In this paper, we propose a fundamentally new way of looking at the
problem of generating referring expressions: We take a referring
expression describing an individual $a$ to be a concept of some
description logic (DL) which is satisfied by $a$ and no other
individual in the universe.  This allows us to organize existing
approaches to GRE in an expressiveness hierarchy.  For instance, the
classical Dale \& Reiter algorithms compute concepts in a description
logic that consists only of conjunctions of atoms, whereas Dale \&
Haddock and other relational algorithms compute concepts in the logic
\el, which permits conjunction and existential quantification.

Furthermore, we show how to apply algorithms for computing the
\emph{bisimulation classes} of a DL model to GRE.  Roughly speaking,
bisimulation classes are sets of individuals that satisfy exactly the
same concepts; if an individual is alone in its bisimulation class,
then the concept corresponding to this class represents a uniquely
referring expression.  We present bisimulation classes algorithms for
the description logics \el\ (only conjunction and existential
quantification) and \alc\ (full modal logic).  By essentially
computing referring expressions for all individuals in the domain at
once, these algorithms systematically avoid the infinite regress
problem.  The \el\ algorithm, which generates the same class of
descriptions as Dale \& Haddock, is capable of generating 67\% of the
relational REs in the
\newcite{viethen06:_algor_for_gener_refer_expres} dataset correctly,
and takes about 400 ms to compute relational REs for all 100
individuals in a random scene.

The paper is structured as follows.  In Section~\ref{sec:bisim}, we
will first define description logic and fundamental concepts such as
bisimulation.  We will then show how to compute bisimulation classes
in Section~\ref{sec:gre} and apply these algorithms to GRE.  In
Section~\ref{sec:discussion}, we evaluate our algorithms and discuss
these results.  Section~\ref{sec:related} compares our approach to
related research; in particular, it shows how various prominent GRE
algorithms can be seen as special cases of ours.
Section~\ref{sec:conclusion} concludes and points to future work.

### 2008-03-11 trunk/Papers/dl-gre-08/related.tex
wrote first versions of intro, related work, efficiency

\section{Introduction} \label{sec:introduction}

\cite{Dale1995}
\cite{dale91:_gener_refer_expres_invol_relat}

>>>>

\section{Related Work} \label{sec:related}

\begin{figure}
  \centering
  \begin{tabular}{l|l}
    GRE algorithm & DL variant \\ \hline
    \newcite{Dale1995} & conjunctions of atoms \\
    \newcite{deemter01:_gener_refer_expres} & propositional logic \\
    \newcite{dale91:_gener_refer_expres_invol_relat} & \el \\
    \newcite{Krahmer2003} & \el \\
    \newcite{kelleher06:_increm_gener_of_spatial_refer} & \el \\
    \newcite{gardent02:_gener_minim_defin_descr} & \alc
  \end{tabular}
  \caption{DL variants used by different GRE algorithms.}
  \label{fig:related}
\end{figure}

The view of GRE as a problem of computing DL concepts allows us to
organize existing approaches to GRE with respect to the logical
connectives of DL which they can use in the referring expressions they
generate.  This is summarized for some approaches in
Fig.~\ref{fig:related}.  At the most unexpressive end of the spectrum,
\newcite{Dale1995} only generate referring expressions that correspond
to conjunctions of atomic concepts.
\newcite{deemter01:_gener_refer_expres} adds the other propositional
connectives (negation and disjunction) to this, whereas
\newcite{dale91:_gener_refer_expres_invol_relat} and its successors
\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} add
existential quantifiers.  \newcite{gardent02:_gener_minim_defin_descr}
combines these strands into an algorithm which computes arbitrary
concepts in \alc. \todo{Is this really what Claire does?}

In this paper, we have presented two algorithms for concepts in \el\
and \alc.  These algorithms compute referring expressions for all
individuals in the domain at once, and thus there is no danger of
infinite regress.  Other approaches for \el\ and higher logics
typically have to invest some effort into avoiding infinite regress --
e.g.\ by requiring that no edge may be used twice
\cite{dale91:_gener_refer_expres_invol_relat} or that no node may be
used twice \cite{kelleher06:_increm_gener_of_spatial_refer}.  The
graph-based algorithm by \newcite{Krahmer2003} avoids infinite regress
with similar ease as our algorithms -- adding an edge to a subgraph
twice doesn't change the subgraph --, but this comes at the price of
NP-complete complexity.

Because they compute referring expressions for all individuals in the
domain together, our algorithms will be strongest in static settings,
such as the generation of descriptions for museum exhibits, in which
the individuals and their properties don't change much.  Nevertheless,
they remain fast enough for real-time use in dynamic settings.
Another advantage that recent algorithms for relational REs
\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} have over
ours is that they can take weights and salience into account.  How
this could be incorporated into our algorithms is an interesting
question for future research.
